\section{Learning Objectives}

This chapter introduces the mechanics of evaluating a trading strategy on
historical data. After completing it you can:

\begin{itemize}
    \item Explain what a backtest is --- and what it is not
    \item Generate trading signals from indicators with no look-ahead bias
    \item Account for transaction costs in P\&L
    \item Evaluate a strategy with the risk-adjusted metrics from
      \cref{ch:returns}
    \item Avoid the three classic backtest traps: overfitting, look-ahead
      bias, and survivorship bias
\end{itemize}

\begin{keyinsight}
A backtest is a \emph{simulation}, not a prediction. It tells you how a
strategy \emph{would have} performed, not how it \emph{will} perform. The number
one failure mode is overfitting to historical data: a strategy that looks
great in backtest and loses money in production.
\end{keyinsight}

\section{What Is Backtesting?}

Backtesting is the replay of a trading strategy over historical price data,
bar by bar, to estimate its hypothetical performance. The engine maintains
state (position, cash, equity) and lets the strategy decide, at each bar,
what action to take given the price history it has seen so far.

\marginnote{%
\textbf{Simulation vs prediction}\\[0.3em]
A backtest does not predict. It is a \emph{counterfactual}: ``if you had
traded this strategy with these rules in this market, here is what would have
happened.'' Any conclusion about the future requires additional assumptions
(market stationarity, capacity, regime persistence) that the backtest itself
cannot validate.
}

Three assumptions are implicit and worth naming:

\begin{enumerate}
    \item \textbf{Market stationarity}: the statistical properties of the
      historical sample are representative of the future. Often false.
    \item \textbf{Capacity}: the strategy can take the simulated sizes without
      moving the market. True for retail; false for institutional size.
    \item \textbf{Execution}: the simulated fills are achievable. Slippage,
      partial fills, and market impact are real and cost money.
\end{enumerate}

\section{Signals and Positions}

A \emph{signal} is a discrete instruction for the current bar. We use three:

\begin{lstlisting}[language=Rust]
#[derive(Debug, Clone, Copy, PartialEq, Eq)]
pub enum Signal {
    Buy,  // Enter long
    Sell, // Exit long (long-only for this chapter)
    Hold, // Do nothing
}
\end{lstlisting}

A \emph{position} is what the engine is currently holding:

\begin{lstlisting}[language=Rust]
#[derive(Debug, Clone, Copy, PartialEq, Eq)]
pub enum Position {
    Long,  // Holding the asset
    Short, // Reserved for later phases
    Flat,  // No position
}
\end{lstlisting}

\begin{keyinsight}
Why separate \texttt{Signal} and \texttt{Position}? A signal is an
\emph{intention}; a position is a \emph{state}. The engine translates
intentions into state transitions, charging transaction costs on every
change. This separation lets the strategy stay declarative (``I want to be
long now'') and the engine stay responsible for the messy realities
(costs, fill assumptions, position limits).
\end{keyinsight}

\section{The Strategy Trait}

\marginnote{%
\textbf{No look-ahead}\\[0.3em]
\texttt{signal(data, index)} receives \texttt{data[..=index]}. The strategy
must never read past the current bar. The engine enforces this by passing a
slice and the current index --- implementations physically cannot access
\texttt{data[index+1..]} without unsafe code.
}

The extension point of the engine is the \texttt{Strategy} trait:

\begin{lstlisting}[language=Rust]
pub trait Strategy {
    fn signal(&self, data: &[Ohlcv], index: usize) -> Signal;
    fn name(&self) -> &str;
}
\end{lstlisting}

The engine is generic over \texttt{S: Strategy}, so new strategies drop in
without engine changes. The trait takes the full slice plus the current index
so the strategy can compute any indicator that fits in the observed history.

\section{SMA Crossover}

The canonical first strategy: go long when a short moving average crosses
above a long moving average (the \emph{golden cross}); exit when it crosses
below (the \emph{death cross}).

\marginnote{%
\textbf{Golden / death cross}\\[0.3em]
$\text{SMA}_S$ crosses above $\text{SMA}_L$ $\Rightarrow$ Buy\\[0.3em]
$\text{SMA}_S$ crosses below $\text{SMA}_L$ $\Rightarrow$ Sell\\[0.5em]
A crossover is a \emph{change in the relationship}, not a level. The previous
bar must satisfy the opposite inequality.
}

\begin{lstlisting}[language=Rust]
pub struct SmaCrossover {
    short_period: usize,
    long_period: usize,
}

impl Strategy for SmaCrossover {
    fn signal(&self, data: &[Ohlcv], index: usize) -> Signal {
        if index < self.long_period { return Signal::Hold; }

        let short_now = sma_at(data, index, self.short_period)?;
        let long_now  = sma_at(data, index, self.long_period)?;
        let short_prev = sma_at(data, index - 1, self.short_period)?;
        let long_prev  = sma_at(data, index - 1, self.long_period)?;

        let crossed_up   = short_prev <= long_prev  && short_now >  long_now;
        let crossed_down = short_prev >= long_prev  && short_now <  long_now;

        if crossed_up   { Signal::Buy  }
        else if crossed_down { Signal::Sell }
        else { Signal::Hold }
    }
    fn name(&self) -> &str { "SMA Crossover" }
}
\end{lstlisting}

\begin{keyinsight}
Crossover strategies are \emph{trend followers}: they only make money when
prices trend. In choppy, mean-reverting markets they lose, because each
crossover is a false signal that gets reversed at the next crossover --- with
transaction costs paid on both sides.
\end{keyinsight}

\section{P\&L with Transaction Costs}

\marginnote{%
\textbf{Transaction cost}\\[0.3em]
At entry: $c_{\text{entry}} = \text{notional} \cdot \tau$\\[0.3em]
At exit: $c_{\text{exit}} = \text{shares} \cdot P_{\text{exit}} \cdot \tau$\\[0.5em]
Net P\&L: $\text{shares} \cdot (P_{\text{exit}} - P_{\text{entry}}) - c_{\text{entry}} - c_{\text{exit}}$
}

Every state change costs money. The engine charges a proportional cost
$\tau$ (e.g.\ $\tau = 0.001$ for 10 basis points) on the traded notional:

\begin{lstlisting}[language=Rust]
// Enter long: spend all cash at the close, paying the cost first.
let notional = capital;
let cost = notional * config.transaction_cost;
shares = (notional - cost) / price;
total_costs += cost;

// Exit long: sell all shares at the close.
let notional = shares * price;
let cost = notional * config.transaction_cost;
total_costs += cost;
capital = notional - cost;
\end{lstlisting}

At each bar, equity is marked to market: \texttt{shares * close} when long,
cash when flat. The per-bar simple returns of the equity curve become the
\texttt{daily\_returns} used by the risk metrics from \cref{ch:returns}.

\section{Performance Evaluation}

\texttt{BacktestResult} delegates to \texttt{qf-04-returns} for the four
headline metrics:

\begin{lstlisting}[language=Rust]
impl BacktestResult {
    pub fn sharpe(&self, rf: f64) -> f64 {
        qf_04_returns::annualized_sharpe(&self.daily_returns, rf, 252.0)
    }
    pub fn sortino(&self, rf: f64) -> f64 {
        qf_04_returns::sortino_ratio(&self.daily_returns, rf)
    }
    pub fn max_drawdown(&self) -> f64 {
        qf_04_returns::max_drawdown(&self.equity_curve)
    }
    pub fn volatility(&self) -> f64 {
        qf_04_returns::annualized_volatility(&self.daily_returns, 252.0)
    }
}
\end{lstlisting}

\begin{keyinsight}
The risk metrics operate on the \emph{equity curve}, not the raw price
series. A strategy that goes flat for half the period has a different risk
profile from buy-and-hold even on the same underlying, because the equity
curve is flat whenever the position is flat. Volatility and drawdown reflect
this exposure pattern automatically.
\end{keyinsight}

\section{Buy-and-Hold Benchmark}

Every active strategy must be compared against the simplest alternative: buy
on the first bar and hold forever. \texttt{BuyAndHold} is a two-line strategy:

\begin{lstlisting}[language=Rust]
impl Strategy for BuyAndHold {
    fn signal(&self, _data: &[Ohlcv], index: usize) -> Signal {
        if index == 0 { Signal::Buy } else { Signal::Hold }
    }
    fn name(&self) -> &str { "Buy and Hold" }
}
\end{lstlisting}

If an active strategy cannot beat buy-and-hold after costs, the active
component is adding noise and cost, not value. The benchmark also lets you
decompose returns: alpha (skill) vs beta (market exposure).

\section{Common Pitfalls}

\begin{description}
    \item[Overfitting] Tuning the SMA periods $(S, L)$ to maximise in-sample
      Sharpe almost always produces a strategy that fails out-of-sample. Use
      walk-forward validation (deferred to a later phase) and be suspicious of
      parameters tuned on the same data you evaluate on.
    \item[Look-ahead bias] Using information not available at decision time.
      Common sources: using the close of bar $t$ to decide at bar $t$ (only
      valid if the trade executes at the close), indicator warm-up that reads
      the future, corporate actions applied with hindsight. Our
      \texttt{signal(data, index)} interface rules out most of these by
      construction.
    \item[Survivorship bias] Backtesting only on companies that survived to the
      present. Companies that went bankrupt are absent, inflating historical
      returns. Use point-in-time universes when available.
\end{description}

\section{Example Output}

\begin{lstlisting}[language=bash]
$ cargo run -p qf-05-backtest --example backtest_demo
==============================
Backtest Results
==============================

Period: 2024-001 to 2024-252

--- Strategy: SMA Crossover (10/30) ---
  Total Return:   3.49%
  Sharpe (ann.):   2.55
  Sortino:         0.31
  Max Drawdown:    -0.41%
  Trades:          1

--- Strategy: Buy and Hold ---
  Total Return:    4.54%
  Sharpe (ann.):   1.70
  Max Drawdown:    -6.35%

Outperformance vs Buy & Hold: -1.04%
\end{lstlisting}

The SMA crossover underperforms buy-and-hold in this synthetic regime: the
cost of false signals in choppy sections outweighs the benefit of avoiding
the mid-year drawdown. This is the typical result for trend-following
strategies in noisy markets, and it is the right lesson to take from a first
backtest.

\section{Exercises}

\begin{enumerate}
    \item \textbf{Cost sensitivity}: Run the 10/30 crossover on the synthetic
      series for $\tau \in \{0, 0.001, 0.0025, 0.005, 0.01\}$. Plot total
      return vs $\tau$. At what cost does the strategy stop being profitable?

    \item \textbf{Parameter sensitivity}: For $\tau = 0.001$, sweep
      $(S, L) \in \{5, 10, 20\} \times \{30, 50, 100\}$ with $S < L$. Which
      configuration maximises Sharpe? Is it the same one that maximises total
      return? Why or why not?

    \item \textbf{EMA crossover}: Implement an EMA crossover strategy by
      replacing SMA with \func{qf\_03\_stocks::ema}. Compare its turnover
      (number of trades) and Sharpe against the SMA crossover on the same
      data.

    \item \textbf{Position sizing}: Modify the engine to invest a fixed
      fraction $f \in (0, 1]$ of capital on each entry instead of 100\%.
      How does the equity curve change for $f = 0.5$? What is the trade-off?

    \item \textbf{Walk-forward (conceptual)}: Split the 252-bar series into
      an in-sample window (first 180 bars) and an out-of-sample window (last
      72 bars). Tune $(S, L)$ on the in-sample, freeze it, and evaluate on
      out-of-sample. How much does Sharpe degrade from in-sample to
      out-of-sample? This is the simplest walk-forward analysis.
\end{enumerate}